\Title

### 2013-01-28 synmut.tex

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines).

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because there evolution can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
however, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011}.
In addition to maintaining protein function and avoiding immune
recognition, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
a skewed codon usage bias towards more or less abundant tRNAs
replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected, but it seems reasonable
that the bulk of positive selection through the immune system be restricted to
amino acid sequences.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation.
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction).

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
synonymous mutations reaching high frequency, very few fix
(panel~\ref{fig:aftsyn}); however, quite some nonsynonymous mutations reach
fixation (panel~\ref{fig:aftnonsyn}).
These observations are quantified in \FIG{fixp}, which stratifies the data
of 7 (resp. 10) patients according to the frequency at which
different mutations are observed (see methods). Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly two years and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$
in panel~\ref{fig:fixp2} (red line). The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back
below.

@@

These observations suggest that many of the synonymous polymorphisms in the part
of \env~that includes the hypervariable regions are deleterious, while outside
this regions polymorphisms are mostly roughly neutral. Note that this does not imply that all 
synonymous mutations are neutral -- only those mutations observed at high frequencies tend 
to be neutral.

@@

One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
secondary structures are relevant for HIV replication, mutations in nucleotides
involved in those base pairs are expected to be deleterious and to revert
preferentially. Many functionally important secondary structure elements have
been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} and the 5'
UTR pseudoknot interacting with the host
tRNA$^\text{Lys3}$~\citep{barat_interaction_1991, paillart_vitro_2002}. It has
been suggested early on that parts of the viral genome that has the potential to
form stems is better conserved than the
remainder~\citep{forsdyke_reciprocal_1995}.

@@

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, any evidence for a contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, within a single patient, we do not observe any bias
towards more human-like CUB in the synonymous mutations that reach fixation
rather than extinction. Second, although codon-optimized HIV seems to perform
better {\it in vitro}, the distance in CUB between HIV and human genes is not
shrinking at the macroevolutionary level. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
\label{fig:SHAPEA}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
in vitro} chemical attack and shown that some nucleotides are more likely to
be involved in RNA secondary folds. C1-C5 regions, in particular, show
conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
among all derived alleles in those regions reaching frequencies of order one,
there is a negative correlation between fixation and involvement in a base
pairing in a RNA stem (left panel). The rest of the genome does not show any
correlation (right panel). There might be too few silent polymorphisms in the
first place, or the signal might be masked by non-functional RNA
structures. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep[genetic hitchhiking,][]{smith_hitch-hiking_1974}.

The potential for hitchhiking is already apparent from the allele frequency
trajectories in \FIG{aft}, where many mutations appear to change rapidly in
frequency as a flock. Deleterious synonymous mutations can be amplified
exponentially by selection on linked nonsynonymous sites, a process known as
{\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
to be advected to high frequency by a linked adaptive mutation, the deleterious
effect of the mutation has to be substantially smaller than the adaptive effect.
The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
effect size of the synonymous alleles (in HIV data is about $-0.2$).
This parameter also reduces synonymous
diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
normalized by the number of possible mutations (first inset).
(Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity.}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we perform computer simulations of
evolving viral populations under selection and rare recombination. For this
purpose, we use the recently published package FFPopSim, which includes a module
dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
pick many random combinations of the following four parameters: effect
size and rate of appearance of beneficial escape mutations, and effect size
and fraction of deleterious synonymous mutations.

The first result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and result in a dependence of the
fixation probability on initial frequency that is compatible with observations
(see \FIG{simfixpvar}).
The fixation probability of synonymous alleles decreases from the neutral
expectation to zero as their fitness effect worsens (\FIG{simfixpvar},
from $-0.0003$ in blue to $-0.003$ in red). Contextually,
the number of alleles that reach high frequencies, which is a proxy for
synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).

To assess the reduction in fixation probability quantitatively, we calculathe
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
fixed). The HIV data correspond to an area of approximately $-0.2$ for synonymous
changes, and essentially zero for nonsynonymous changes.

Among all simulations, we select the one that show a large depression in
fixation probability of synonymous mutations but, simulteneously, a high
synonymous diversity, as observed in HIV data. This selection imposes a strong
constraint on the parameters, in particular on the ones related to synonymous
changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \lesssim 0.002$, is required. This result fits well
the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
Moreover, it corresponds to {\it a priori} theoretical expectations:
(a) since high frequency bins are enriched in neutral mutations (it is easier for
them to hitchhike), the hallmark of deleterious mutations will be visible there
only if they are pervasive; (b) genetic draft stops working if the
deleterious effect size becomes comparable to the adaptive effect size of escape
mutations, because the double mutant has little or no fitness advantage.

As long as only features of synonymous mutations are filtered, as performed
above, no strong correlation of the two nonsynonymous parameters, effect size
and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
shown possess a variety of shapes and colors. In most of these simulations, 
however, nonsynonymous mutations almost always fix once they reach high frequencies
-- their nonsynonymous fixation area is well above zero. This stands in contrast
with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
were neutral. Although there are no doubts that nonsynonymous variation in the
variable regions is driven by positive selection and not by genetic drift, the
biological reason for the neutral-like curve in \FIG{fixp} is unclear.
Inspecting the trajectories of nonsynonymous mutations
suggests the rapid rise and fall of many alleles. We test two possible
mechanisms that are biologically plausible and could explain the transient rise
of nonsynonymous mutations: time-dependent selection and within-epitope
competition.

The former explanation can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average time needed by an escape mutant to cross frequencies of order one.
In the alternative explanation, several different escape
mutations within the same epitope might arise almost simultaneously and start to
spread. Their fitness benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency rapidly, while the one with the smallest cost in terms of replication,
packaging, etc. is most likely to eventually fix. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
We tested both hypotheses in our simulation framework and they both seem to
achieve a reduced, neutral-like fixation probability while maintaining a high
rate of substitutions, compatibly with HIV measurements. In real HIV infections,
both mechanisms are likely to be playing a role.

>>>>

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines). Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because there evolution can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
however, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011}.
In addition to maintaining protein function and avoiding immune
recognition, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
a skewed codon usage bias towards more or less abundant tRNAs
replicate better or worse \comment{replicate better or produce more protein?},
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but it seems reasonable
that the bulk of positive selection through the immune system be restricted to amino acid sequences.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction).

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
synonymous mutations reaching high frequency, very few fix
(panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).
These observations are quantified in \FIG{fixp}, which stratifies the data
of 7 (resp.~10) patients according to the frequency at which
different mutations are observed (see methods). Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$
in panel~\ref{fig:fixp2} (red line). The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back
below.

@@

These observations suggest that many of the synonymous polymorphisms at
intermediate frequencies in the part of \env~that includes the hypervariable
regions are deleterious, while outside this regions polymorphisms are mostly
roughly neutral. Note that this does not imply that all synonymous mutations are
neutral -- only those mutations observed at high frequencies tend to be neutral.

@@

One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. It has
been suggested early on that parts of the viral genome that has the potential to
form stems is better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

@@

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
and -pessimized viruses have recently been generated and shown to replicate
better or worse than wild type strains,
respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
coleman_virus_2008}. We do not find, however, any evidence for a contribution of
CUB to the ultimate fate of synonymous alleles. Several lines of thought support
this result. First of all, within a single patient, we do not observe any bias
towards more human-like CUB in the synonymous mutations that reach fixation
rather than extinction. Second, although codon-optimized HIV seems to perform
better {\it in vitro}, the distance in CUB between HIV and human genes is not
shrinking at the macroevolutionary level. Third, it is a common phenomenon for
retroviruses to use variously different codons from their hosts, and CUB effects
on fitness are thought to be so small that divergent nucleotide composition has
been suggested as a possible mechanism for viral
speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
not very different from other parts of the HIV genome, whereas the reduced
fixation probability is only observed there. In conclusion, although we cannot
exclude an effect of CUB on fitness as a general rule, we expect it to be a
minor effect in our context. \comment{Need to read literature}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
\label{fig:SHAPEA}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities. \comment{Is this C2-V5 only?} 
Panel A) Synonymous substitions in the C2-V5 region have substantially higher
SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost. This difference is quantified by a cumulative histogram
($p=0.002$, Kolmogorov-Smirnov two sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B shows the fixation probablility of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep[genetic hitchhiking,][]{smith_hitch-hiking_1974}.

The potential for hitchhiking is already apparent from the allele frequency
trajectories in \FIG{aft}, where many mutations appear to change rapidly in
frequency as a flock. Deleterious synonymous mutations can be brought to high
frequency by selection on linked nonsynonymous sites, a process
known as {\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
to be advected to high frequency by a linked adaptive mutation, the deleterious
effect of the mutation has to be substantially smaller than the adaptive effect.
The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, that shows the distribution of times for synonymous
alleles to reach the fix or get lost starting from intermediate frequencies. The
typical time to loss is of the order of 500 days. If this loss is driven by the
deleterious effect of the mutation, this corresponds to deleterious effects of
roughly $s_d \sim - 0.002$ per day.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Evolutionary parameters compatible with observations.
Panel A) The depression in $P_\text{fix}$ depends on the deleterious
effect size of the synonymous alleles (in HIV data is about $-0.2$).
Increasing this parameter also reduces synonymous
diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
normalized by the number of possible mutations (first inset).
Panel B) To assess the parameter space compatible with observations, we run many
simulations with random parameters for deleterious effect size, fraction of
deleterious synonymous sites, average size of escape mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. \comment{should we extend the
parameters space maybe to $10^{-4}$?}}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we perform computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}.

The first result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and results in a dependence of the
fixation probability on initial frequency that is compatible with observations
(see \FIG{simfixpvar}).
The fixation probability of synonymous alleles decreases from the neutral
expectation to zero as their deleterious effect increases
(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red). Along with the
reduction in fixation probability, the number of alleles that reach high
frequencies, which is a proxy for synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).

To assess the reduction in fixation probability quantitatively, we calculate
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
fixed). The HIV data correspond to an area of approximately $-0.2$ for synonymous
changes, and essentially zero for nonsynonymous changes.

We pick many random combinations of the following four parameters: effect size
and rate of appearance of beneficial escape mutations, and effect size and
fraction of deleterious synonymous mutations.
Among all simulations, we select the one that show a large depression in
fixation probability (>-0.??) of synonymous mutations but, simulteneously, a
high synonymous diversity (>0.0? per site), as observed in HIV data. This
selection imposes a strong constraint on the parameters, in particular on the
ones related to synonymous changes, i.e.~their deleterious fraction and effect
size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \lesssim 0.002$, is required. This result fits well the
expectation based on the fixation/extinction times above (see \FIG{fixp1}).
Moreover, it corresponds to {\it a priori} theoretical expectations:
(a) since high frequency bins are enriched in neutral mutations (it is easier
for them to hitchhike), the hallmark of deleterious mutations will be visible
there only if they are pervasive; (b) genetic draft stops working if the
deleterious effect size becomes comparable to the adaptive effect size of escape
mutations, because the double mutant has little or no fitness advantage.

As long as only features of synonymous mutations are filtered, as performed
above, no strong correlation of the two nonsynonymous parameters, effect size
and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
shown possess a variety of shapes and colors. In most of these simulations, 
however, nonsynonymous mutations almost always fix once they reach high frequencies
-- their nonsynonymous fixation area is well above zero. This is incompatible
with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
were neutral. Although there are no doubts that nonsynonymous variation in the
variable regions is driven by positive selection and not by genetic drift, the
biological reason for the neutral-like curve in \FIG{fixp} is unclear.
Inspecting the trajectories of nonsynonymous mutations
suggests the rapid rise and fall of many alleles. We test two possible
mechanisms that are biologically plausible and could explain the transient rise
of nonsynonymous mutations: time-dependent selection and within-epitope
competition.

The former explanation can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average time needed by an escape mutant to cross frequencies of order one.
In the alternative explanation, several different escape
mutations within the same epitope might arise almost simultaneously and start to
spread. Their benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency rapidly, while the one with the smallest cost in terms of replication,
packaging, etc.~is most likely to eventually fix. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
We tested both hypotheses in our simulation framework and they both seem to
achieve a reduced, neutral-like fixation probability while maintaining a high
rate of substitutions, compatibly with HIV measurements. In real HIV infections,
both mechanisms are likely to be playing a role.
